---
title: "NCBI GEO Metadata Survey"
author: "Sean Davis"
date: "April 28, 2015"
output: 
  html_document:
    theme: readable
    highlight: pygments
---

## Get metadata sqlite file

```{r echo=FALSE,result="hide"}
library(knitr)
library(pander)
opts_chunk$set(warning=FALSE,message=FALSE,fig.width=8,fig.height=8)
```


```{r getsqlite}
library(GEOmetadb)
sfile = 'GEOmetadb.sqlite'
if(!file.exists(sfile)) {
  sfile = getSQLiteFile() 
}
```

## Set up dplyr stuff....

Just for fun, instead of using SQL queries, we can rely on dplyr to do the 
SQL for us.

```{r multiplatform}
library(dplyr)
gmdb = src_sqlite(sfile)
# List available tables in the database
src_tbls(gmdb)
tgse = tbl(gmdb,'gse')
tgsm = tbl(gmdb,'gsm')
tgpl = tbl(gmdb,'gpl')
```

## Queries of interest

### Simple table summaries

```{r results='markup'}
tablestats = data.frame(entity=c('Series','Samples','Platforms'),
                        Count=c(nrow(tgse),nrow(tgsm),nrow(tgpl)))
pander(tablestats,justify=c('left','right'))
```

### Multiplatform datasets by gpl

One way of measuring multiplatform GEO Series in GEO is to map the GSE
identifiers to their associated GPL identifiers.  This probably overestimates
the actual complexity of the data, as some GEO series map to multiple arrays
of the same technology.  For example, there is a GEO Series, GSE1097, that
has 153 platforms, but this was a set of whole-genome tiling arrays.

```{r multiplatformGPL}
tgse_gpl = tbl(gmdb,'gse_gpl')
gse_gpl_count = select(tgse_gpl,gse) %>% 
  group_by(gse) %>%
  summarise(count=n()) %>%
  filter(count>1) %>%
  collect()
# and make a cut base on number of platforms associated with a GSE
cut_count = cut(gse_gpl_count$count,
                breaks=c(2,3,5,10,20,max(gse_gpl_count$count)))
table(cut_count)
```

And a small barplot with the binned number of GPLs per GSE:

```{r multiplatformGPLPlot}
barplot(table(cut_count),
        main="GEO Series by Number of GEO Platforms",
        sub="Includes only GEO Series with more than one platform")
```

### Multiplatform based on assay type

Another way to quantify "multiplatform" is to look at the different assay 
types in a GSE.  This may underestimate the number somewhat, but it is 
probably closer to what we want to capture.  The `type` column in the GSE
table includes a semicolon-separated list of experiment types. While perhaps
not a perfect list, it at least allows us to quantify the number of 
experiments of each type. The number of each `type` is given in the next table.

```{r GSETypes}
library(tidyr)
gse_type = select(tgse,gse,type) %>%
  transform(type = strsplit(type,';\\t')) %>%
  unnest(type) 
type_count = select(gse_type,type) %>%
  group_by(type) %>%
  summarize(count=n()) %>% 
  arrange(desc(count))
pander(type_count,justify=c('left','right'))
```

We can summarize the number of different types per GEO Series.  The results are given in the next table.

```{r multiplatformByGSEType}
gse_by_type_count = select(gse_type,gse) %>%
  group_by(gse) %>%
  summarize(NumberOfDataTypes=n()) %>%
  group_by(NumberOfDataTypes) %>% 
  summarize(NumberOfGEOSeries=n())
pander(gse_by_type_count,justify=c('right','right'))
```

## Data growth over time

I have summarized the data by profiling type and year of submission.  To
keep the data clean, I have included only the 6 most common GSE types
for the plot.  Feel free to adjust.

```{r gseByDateAndType}
library(lubridate)
gse_type_year = select(tgse,gse,type,year=submission_date) %>%
  transform(year=year(as.Date(year))) %>%
  filter(year<2015) %>%
  transform(type = strsplit(type,';\\t')) %>%
  unnest(type) %>% 
  right_join(type_count[1:6,'type']) %>%
  group_by(type,year) %>%
  summarize(submissions=n())
```

The following plot shows submissions per year (not cumulative number of
submissions) cut off at 2014. There are clearly trends toward sequencing
over microarrays, but microarray expression profiling still greatly 
exceeds sequencing--note y-scale is log10.

```{r submissionsPlot}
library(ggplot2)
ggplot(gse_type_year,aes(x=year,y=submissions,group=type,color=type)) +
  geom_line() + 
  scale_y_log10() + 
  theme(legend.position="bottom") + 
  guides(colour = guide_legend(nrow = 6))
```

## sessionInfo()

```{r sessionInfo}
sessionInfo()
```
